---
title: "Visualizing green power market size and consumer choice by state using ggplot2 in R"
author: "Uni Lee"
date: "9/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(tidyr)
library(maps)
library(ggplot2)
library(janitor) # To clean column names
library(ggrepel) # For labeling state names
library(readxl)
library(ggpubr) # For putting plots together
library(rmarkdown)
```

# Mapping electricity market structures by state

Not every state has an open electricity market, where consumers can freely choose their electricity supplier. [Electric Choice](https://www.electricchoice.com/map-deregulated-energy-markets/) categorizes states into deregulated and regulated markets. Using the ggplot2 and maps package in R, I will visualize availability of consumer choice for electricity by state.

```{r}
# Import electricity market structure data from Electric Choice (source: https://www.electricchoice.com/map-deregulated-energy-markets/).
markets_raw <-  read.csv("Data/power_markets_us.csv")
markets <- markets_raw %>% 
  clean_names() %>%
  select(state, electric) %>%
  mutate(electric = case_when(
            electric == "No" ~ "Regulated",
            electric == "Yes" ~ "Deregulated",
            electric == "Yes*" ~ "Partially regulated")) # Treat California as a regulated market
  
markets <- markets %>% 
  mutate(electric=ifelse(markets$state %in% c("Washington DC", "Washington"), "Mixed", markets$electric)) %>% # Treat Washington as a mixed market
  drop_na() 


# Import US geography data to a dataframe for plotting 
us_states <- map_data("state")

# Merge market data and the geography by "region". 
## First, we have to make sure that the values in the "region" column from two datasets match. We will lower all characters to do that. 
markets$state <- tolower(markets$state)
markets <- markets %>% rename(region = state)
## Adding regional abbreviation
us_abb <- read.csv("Data/us_abb.csv") %>% rename(
  region=State.Name, abb = USPS.Abbreviation) 
us_abb$region <- tolower(us_abb$region)

markets <- left_join(markets, us_abb, by="region")

# Merging geographical and electricity choice datasets
choice_data <- left_join(us_states, markets, by="region") %>% drop_na(electric)

rmarkdown::paged_table(choice_data)

# Create a dataframe for labeling 
label_dt <- choice_data %>% group_by(region) %>%
  summarize(long=mean(range(long)), lat=mean(range(lat))) %>% na.omit() ## position of the labels are determined by taking the mean of long and lat. 
label_dt <- left_join(label_dt, markets, by="region") 

# Plot
choice_map <- ggplot(data= choice_data,
       aes(x=long, y=lat, group=group,
           fill=electric)) + 
  geom_polygon(size=0.1) + 
  geom_path(size=0.03) +  
  geom_text(data=label_dt,
                   aes(x=long, y=lat, group=region, label=abb),
                   size=2, hjust=1) + 
  coord_map(projection="albers", lat0=39, lat1=45) + 
  labs(title="Availability of consumer choice for green power by state") + 
  theme_classic() + 
  theme(legend.position = "bottom",
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank()) 
choice_map 
```

The differences in structures of the electricity market is important because they affect how electricity is procured and the degree to which consumers can purchase low-carbon electricity. 

A third of states in the US have *regulated* electricity markets, where utility companies operate as vertically integrated monopolies.  In regulated markets, retail electricity prices are overseen by state regulators, protecting consumers from high prices. The downside is that consumers can't choose their supplier. If the utility offers green pricing or green tariff programs, then consumers can opt for renewable electricity. If not, consumers don't have any other choice.  

In *deregulated* markets, retail and wholesale markets for electricity are open for competition to independent power producers, while transmission systems are operated by independent system operators or regional transmission organizations (ISOs, RTOs). In deregulated markets, retail and wholesale electricity prices are determined by market forces. Here, energy consumers can choose their supplier according to their preference. For example, consumers can opt for a supplier that sources electricity only from renewable sources at different prices. 

Some markets are *partially regulated*. For example, in California, wholesale markets are deregulated, while the retail market remains open only for a small number of consumers who opt for renewable energy through [direct access program](https://www.cpuc.ca.gov/industries-and-topics/electrical-energy/electric-costs/learn-more-about-costs-and-rates) or ([community choice aggregation](https://en.wikipedia.org/wiki/Community_Choice_Aggregation)). 


# Voluntary green power market size by state

Since consumers have limited choice for renewable electricity, it is plausible to assume that voluntary green power markets are smaller in regulated markets than deregulated ones. After all, the market can grow only if consumers can express their willingness to pay in a free market, right?

To test this assumption, I wrote the following code to draw a heatmap of the voluntary green power market sizes in the US by state. The market size can be expressed in terms of the number of customers and the amount of purchase (MWh), so there are two maps to see. 

Resulting maps show that Texas has the biggest amount of green power electricity, whereas California has highest number of customers who voluntarily purchase green power. 

```{r}
# Load data green power market data from NREL
gp_gen_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=4, skip=4) %>% clean_names()
gp_cus_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=5, skip=4) %>% clean_names()

gp_gen <- gp_gen_raw %>% select(state, tot_gen = total) # market size by generation
gp_cus <- gp_cus_raw %>% select(state, tot_cus = total) # market size by number of customers
gp <- merge(gp_gen, gp_cus, by="state") %>% rename(abb=state)

paged_table(gp)

mkt_data <- left_join(choice_data, gp, by="abb") # Add the new data to the dataset for mapping

options(scipen=999) # Turn off scientific notation in the legend

# Build a function for creating a map to save some code lines
map_gp <- function(data, fill){
  ggplot(data= data,
       aes(x=long, y=lat, group=group)) +  
  geom_polygon(aes(fill=fill)) + # fill goes to geom_polygon for continuous data
  geom_path(size=0.03) +  
  scale_fill_gradientn(colours = rev(heat.colors(10)),na.value = "grey90", 
                       guide = guide_colourbar(barwidth = 25, barheight = 0.4,
                                               title.position = "top")) +
  geom_text(data=label_dt,
                   aes(x=long, y=lat, group=region, label=abb),
                   size=2, hjust=1) + 
  coord_map(projection="albers", lat0=39, lat1=45) + 
  labs(title= "Size of voluntary green power by state (MWh)") + 
  theme_classic() + 
  theme(legend.position = "bottom",
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank()) 
  }

gen_map <- map_gp(data=mkt_data, fill=mkt_data$tot_gen)
cus_map <- map_gp(data=mkt_data, fill=mkt_data$tot_cus)

gen_map
cus_map
```

Looking at these maps, we cannot conclude whether electricity structure affects the size of green power markets. But, we can infer that large-scale energy consumers (i.e. IT companies) participate more actively in the green power market in Texas (deregulated market),  while small-to-medium scale energy consumers are more active in California (partially deregulated market).

References 

Electric Choice. "Deregulated Energy Markets". https://www.electricchoice.com/map-deregulated-energy-markets/

Cleary, K., Palmer, K. (2020). US Electricity Markets 101. Resources for the Future. https://media.rff.org/documents/US_Electricity_Markets_101.pdf