---
title: "Green power"
author: "Uni Lee"
date: "9/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(tidyr)
library(readxl)
library(janitor)
library(ggplot2)
library(ggpubr)
```

# How fast has the voluntary green power market grown in the US? 

What kind of electricity do you use at home? As climate crisis looms over, there is a global movement of companies and citizens supporting energy transition by voluntarily buying clean electricity for their use. The [RE100](https://www.there100.org/) campaign attests to this growing movement, whose member companies commit to using 100% renewable electricity. By 2020, over 300 companies have joined the RE100 campaign.

As the demand for clean electricity rises, the market size of voluntary green power is growing. In this blog, I will demonstrate the growth of this market by exploring the US Voluntary Green Power Market Data from 2010 to 2019, provided by [National Renewable Energy Laboratory](https://data.nrel.gov/submissions/151). _Thank you NREL, for making this wonderful dataset publicly available!_ 


Although the growth of voluntary renewable energy procurement is truly a global phenomenon, I limit the scope of analysis to the US. In the US, electricity market structures vary from state to state. This allows us to control for structural factors when considering the growth of green power.You can read more about different electricity market structures in my previous post [*here*](https://conscioustableblog.wordpress.com/2021/09/26/building-heatmaps-for-green-power-market-by-state-using-ggplot2-in-r/). 

Overall in 2019, the US voluntary green power market was FOUR times bigger than it was in 2010, both in terms of sales and number of customers. This is huge: we are talking about a growth rate of 330%! 

The graph below shows how the market grew in terms of sales. The growth of unbundled REC sales and Power Purchase Agreements (PPA) is outstanding. PPAs are considered to be the most desirable procurement method, since they assure additionality and decrease investment risks for renewable energy producers. Companies can also procure renewable electricity by opting out for green pricing or green tariff programs through the utility company, or purchasing unbundled RECs in the voluntary REC market. 

```{r}
# Load data 
# Source: Heeter, Jenny; O'Shaughnessy, Eric (2020): U.S. Voluntary Green Power Market Data 2019. National Renewable Energy Laboratory. https://data.nrel.gov/submissions/151

gp_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=3, skip=2) %>% clean_names()

# Plot growth of green power market by sales and number of customers
sales <- ggplot(gp_raw, aes(x=factor(year), y= sales_million_m_wh, group=segment, fill=segment)) + 
  geom_area() + 
  theme_classic() +
  theme(legend.position="right",
        axis.title.x = element_blank(),
        legend.title = element_blank()) + 
  labs(y = "Sales in Million MWh",
       title="US Green Power Market Size",
       subtitle="in terms of sales (MWh)") +
  scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                      "#0072B2","#D55E00")) 
sales 
```

The following graph shows the increase in the number of customers. An outstanding growth is seen in Community choice aggregation (CCA), which is available only in fully liberalized markets and California. CCA reflects the rising demand among small-scale energy consumers. That's why their overall sales volume is small. PPAs are not visible on this graph because only large-scale users can opt for this option. 

```{r}
cust <- ggplot(gp_raw, aes(x=factor(year), y= customers_x1000, group=segment, fill=segment)) + 
  geom_area() + 
  theme_classic() +
  theme(legend.position="right",
        axis.title.x = element_blank(),
        legend.title = element_blank()) + 
  labs(y = "Number of customers (x1000)", 
       title="US Green Power Market Size",
       subtitle="in terms of customers") +
  scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                      "#0072B2","#D55E00"))

cust
```


# Do liberalized electricity markets have bigger green power markets? 

As we discussed in the previous blog, electricity market structure affects the degree to which consumers procure renewable electricity. I would assume that fully liberalized power markets have bigger green power markets because consumers have more procurement options.  

To test this assumption, let us group the data by electricity market structure and see if 1) deregulated markets have higher market growth than regulated ones, and 2) which procurement methods are most prominent for each market structure. 

The following code joins electricitiy market structure data from [ Electric Choice](https://www.electricchoice.com/map-deregulated-energy-markets/) with NREL's 2019 green power market data to visualize green power market size by procurement type for different market structures. 

```{r}
# Market structure data
markets_raw <-  read.csv("Data/power_markets_us.csv")
markets <- markets_raw %>% 
  clean_names() %>%
  select(state, electric) %>%
  mutate(electric = case_when(
            electric == "No" ~ "Regulated",
            electric == "Yes" ~ "Deregulated",
            electric == "Yes*" ~ "Partially deregulated")) %>%
  drop_na() %>%
  mutate(state=
           ifelse(state=="Washington DC", "district of columbia", state)) %>%
  mutate(state=tolower(state))

# US State names and abbreviations (built-in vectors in R)
us_abb <- tibble(state.name, state.abb) %>% 
  mutate(state.name=tolower(state.name)) %>%
  rename(state=state.name, abb=state.abb)
dc <- c("district of columbia", "DC") # Add DC to the list
us_abb <- rbind(us_abb, dc)

# Merge them together
markets_abb <- left_join(markets,us_abb, by="state")

# Load NREL voluntary green power data by state and procurement methods
gp_gen_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=4, skip=4) %>%
  rename(abb=State) %>% 
  select(-Total)

gp_gen_long <- gp_gen_raw %>% pivot_longer(
  cols=`Utility Green Pricing`:PPAs,
  names_to="procure_type",
  values_to = "gen_mwh")

gp_cus_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=5, skip=4) %>%
  rename(abb=State) %>% 
  select(-Total)

gp_cus_long <- gp_cus_raw %>% pivot_longer(
  cols=`Utility Green Pricing`:PPAs,
  names_to="procure_type",
  values_to = "num_customer")

# Merge generation and number of customers
options(scipen=999)
gp_all <- left_join(gp_gen_long, gp_cus_long, by=c("abb", "procure_type"))

# Add market structure data
gp_everything <- left_join(gp_all, markets_abb, by="abb")

# Summarize data by market structure
mkt <- gp_everything %>% group_by(electric, procure_type) %>%
  summarise(gen_mwh = sum(gen_mwh, na.rm=TRUE),
            num_customer = sum(num_customer, na.rm=TRUE))

# To order bars
mkt$electric <- factor(mkt$electric, levels=c("Regulated", "Deregulated", "Partially deregulated"))
ggplot(data=mkt, aes(x=electric, y=gen_mwh/1000, fill=procure_type)) +
  geom_bar(stat="identity") + 
    scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                      "#0072B2","#D55E00")) + 
  theme_light() + 
    theme(axis.title.x = element_blank(),
        legend.title = element_blank()) +
  labs(y="GWh",
       title="Green power procurement in 2019",
       subtitle="by market structure and procurement type") 
```

Note that the third bar "Partially deregulated" is the portion of electricity procured in California. Considering the fact that the CCA, PPAs and competitive supplier portion of California's green power was procured in similar ways to deregulated market, they can be considered as a deregulated market. 

Therefore, we can conclude that voluntary green power markets have similar market size in different market structure. However, they differ in terms of how energy consumers procure green power. In regulated markets, most green power is procured through purchasing unbundled RECs. In deregulated markets, mroe consumers tend to opt out for a different supplier in the retail market. 


