---
title: "Global Power Plant Database WRI"
author: "Uni Lee"
date: "11/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) 
library(tidyverse)
library(janitor)
library(readxl)
library(ggmap) # To match geocoordinates to municipalities 
library(revgeo)
library(tidygeocoder)

`!%in%` <- Negate(`%in%`)
options(scipen=999)
```

Gathering high-resolution power capacity data my aggregating power-plant level data from WRI. 

```{r}
ppt <- read.csv("Data/global_power_plant_database_v_1_3/global_power_plant_database.csv") 

ppt_us <- ppt %>% 
  select(country, name, capacity_mw, latitude, longitude, primary_fuel, other_fuel1, other_fuel2, other_fuel3, commissioning_year, owner) %>%
  filter(country=="USA")
ppt_re_us <- ppt_us %>% filter(
  primary_fuel %in% c("Solar","Wind"))

us_solar_wind <- read.csv("Data/us_re.csv") 




#ppt_re_ct <- ppt_re_us %>%
#  mutate(
  #  county = reverse_geocode(ppt_re_us,
   #             latitude,
    #            longitude,
     #           return_input=FALSE,
      #          method="osm",
       #         full_results=TRUE
        #        )$county
    #)
```

Power plants by year of commissioning (= new capacity of that year)

```{r}
# All states and counties
us_re_agg <- us_solar_wind %>% group_by(start_yr, state, county) %>%
  summarise(re_cap = sum(capacity_mw)) 

# CA and CO
us_re_coca <- us_re_agg %>% filter(
  state %in% c("California", "Colorado")
) %>% filter(start_yr > 1999) 

ggplot(us_re_coca, aes(x=start_yr, y=re_cap, fill=state)) +
  geom_bar(stat="identity", position="dodge")
```
=> We can't use this data because it does not include decomissioned power plant data. 
# State RE capacity addition

* EIA. (2021): Existing Nameplate and Net Summer Capacity by Energy Source, Producer Type and State, 1990-2020 . US Energy Information Administration. https://www.eia.gov/electricity/data/state/

```{r}

net_capa_raw <- read.csv("Data/existcapacity_annual.csv", skip=1) %>% clean_names() # in MWh

net_capa_raw$nameplate_capacity_megawatts <- as.numeric(gsub(",","",net_capa_raw$nameplate_capacity_megawatts))

net_capa_tot <- net_capa_raw %>% filter(year>=2000, producer_type=="Total Electric Power Industry",
                                        fuel_source %in% c("Wind", "Solar Thermal and Photovoltaic")) %>% 
  select(year, state_code, fuel_source, capacity = nameplate_capacity_megawatts) %>%
  group_by(year, state_code) %>%
  summarize(capacity=sum(capacity))
net_capa_tot

coca <- net_capa_tot %>% filter(state_code %in% c("CO", "CA"))

ggplot(coca, aes(x=year, y=capacity, color=state_code)) + geom_line() + 
  theme_bw()
```
Calculate annual capacity from powerplant data

```{r}

```

