---
title: "Finding a running variable"
author: "Uni Lee"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(tidyr)
library(readxl)
library(janitor)
library(ggplot2)
library(ggpubr)
```
# State RE capacity addition

* EIA. (2021): Existing Nameplate and Net Summer Capacity by Energy Source, Producer Type and State, 1990-2020 . US Energy Information Administration. https://www.eia.gov/electricity/data/state/

```{r}

net_capa_raw <- read.csv("Data/existcapacity_annual.csv", skip=1) %>% clean_names() # in MWh

net_capa_raw$nameplate_capacity_megawatts <- as.numeric(gsub(",","",net_capa_raw$nameplate_capacity_megawatts))

net_capa_tot <- net_capa_raw %>% filter(year>=2009, producer_type=="Total Electric Power Industry",
                                        fuel_source %in% c("Wind", "Solar Thermal and Photovoltaic")) %>% 
  select(year, state_code, fuel_source, capacity = nameplate_capacity_megawatts) %>%
  group_by(year, state_code) %>%
  summarize(capacity=sum(capacity))
net_capa_tot
```


# State green power record  

```{r}
# Load data 
# Source: Heeter, Jenny; O'Shaughnessy, Eric (2020): U.S. Voluntary Green Power Market Data 2019. National Renewable Energy Laboratory. https://data.nrel.gov/submissions/151

gp_2017 <- read_xlsx("Data/NREL_Green Power 2017 Data.xlsx", sheet=5, skip=2) 

gp_2019 <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=4, skip=2) 

gp_2020 <- read_xlsx("Data/NREL_Green_Power_Data_v2020.xlsx", sheet=4, skip=2) 

# 2018 data doesn't have state-by-state data

# Wrangle
long_2017 <- gp_2017 %>% pivot_longer(cols=`Utility Green Pricing`:PPAs,
                         names_to = "Sourcing",
                         values_to = "MWh") %>%
  mutate(Year=2017) %>% select(State, Year, Sourcing, MWh)


long_2019 <- gp_2019 %>% pivot_longer(cols=`Utility Green Pricing`:PPAs,
                         names_to = "Sourcing",
                         values_to = "MWh") %>%
  mutate(Year=2019) %>% select(State, Year, Sourcing, MWh)

long_2020 <- gp_2020 %>% pivot_longer(cols=`Utility Green Pricing`:PPAs,
                         names_to = "Sourcing",
                         values_to = "MWh") %>%
  mutate(Year=2020) %>% select(State, Year, Sourcing, MWh)

gp_all <- full_join(full_join(long_2017, long_2019), long_2020)

```

# DiD: looking for a state similar to California

NREL's 3010 Renewable Energy Technical Potential implies that Colorado has similar renewable energy (solar and wind combined) potential as California. Until 2010, Colorado and California both had regulated electricity markets. In 2010, Community choice aggregation and Direct Access were introduced in California, while keeping the rest of the market regulated. Colorado remained fully regulated, with no direct PPA records to this day. Consumers can only choose green pricing programs provided by conventional utilities (such as Xcel Energy or Black Hills Energy, or municipal utilities. 

```{r}
co_ca <- gp_all %>% filter(State %in% c("CA", "CO"))

ggplot(co_ca, aes(x=Year, y=MWh)) +
  geom_area(aes(fill=Sourcing)) +
  facet_wrap(~State) + theme_classic() +
  theme(legend.position="bottom")
```
```{r}
co_ca_re_cap <- net_capa_tot %>% filter(state_code %in% c("CA", "CO"))

ggplot(co_ca_re_cap, aes(x=factor(year), y=capacity, group=state_code)) + 
  geom_line(aes(color=state_code)) +  theme_minimal() +
  theme(axis.title.x=element_blank())+
  labs(y="Solar and Wind Capacity in MW")
```


2. Fixed Effects: finding control and treatment groups 

```{r}
direct <- gp_all %>% filter(Sourcing %in% c("PPAs", "CCAs", "Utility Renewable Contracts"))

direct_tot <- direct %>% group_by(State, Year) %>%
  summarize(GP = sum(MWh))

control_sub <- direct_tot %>% filter(GP < 50000)
control_states <- unique(control_sub$State)

`%!in%` <- Negate(`%in%`)
control <- direct_tot %>% filter(State %in% control_states)
treatment <- direct_tot %>% filter(State %!in% control_states)

unique(control$State)
unique(treatment$State)

ggplot(control, aes(x=factor(Year), y=GP)) + geom_point() +
         facet_wrap(~State)
ggplot(treatment, aes(x=factor(Year), y=GP)) + geom_point() +
         facet_wrap(~State)

ggplot(control, aes(x=factor(Year), y=GP, fill=State)) + geom_bar(stat="identity", position="dodge") 


```

