---
title: "Cleaning and merging various datasets on US electricity production"
author: "Uni Lee"
date: "10/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) 
library(tidyverse)
library(janitor)
library(readxl)
`!%in%` <- Negate(`%in%`)
options(scipen=999)
```

# Collecting various datasets on US electricity

I am writing my thesis on green power markets in the United States. My hypothesis is that the green power electricity sales increased the share of renewable energy in the US. To test my hypothesis, I would like to do a multivariate regression analysis, controlling for variables that affect growth of renewables: income,  electricity market structure, cost competitiveness of renewables, renewable energy support schemes.

Of course, there is no perfect dataset that is ready for analysis. So I had to hunt for data from various sources. This is what I've harvested:

* EIA. (2021). Electricity Sales to Ultimate Customers by State by Sector by Provider (EIA-861). US Energy Information Administration]https://www.eia.gov/electricity/data/state/

```{r}
net_sales_raw <- read.csv("Data/sales_annual_bystate.csv", skip=1) %>% clean_names() # in MWh
```

* EIA. (2021): Existing Nameplate and Net Summer Capacity by Energy Source, Producer Type and State, 1990-2020 . US Energy Information Administration. https://www.eia.gov/electricity/data/state/

```{r}
net_capa_raw <- read.csv("Data/existcapacity_annual.csv", skip=1) %>% clean_names() # in MWh
```

* Heeter, Jenny; O'Shaughnessy, Eric (2020): U.S. Voluntary Green Power Market Data 2019. National Renewable Energy Laboratory []. https://data.nrel.gov/submissions/151

```{r}
# Historical data (no state-by-state data)
gp_historical_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=2, skip=2) %>% clean_names()

# Voluntary Green Power Sales (MWh) by State 
gp_2019_gen_raw <- read_xlsx("Data/NREL_Green_Power_Data_v2019.xlsx", sheet=4, skip=4) %>% clean_names()
```

# Calculating the share of green power sales in US electricity sales

The EIA dataset on US electricity generation is broken down into states and energy source. We will summarize this dataset at US level, by year. Then, we merge it with NREL green power dataset to calculate share of green power sales.

```{r}
net_sales_us <- net_sales_raw %>% 
  filter(state=="US", industry_sector_category == "Total Electric Industry", year >= 2010) %>%
  select(year, total) 
         
net_sales_us$total <- as.numeric(gsub(",","",net_sales_us$total))

compare_sales <- merge(net_sales_us, gp_historical_raw, by="year")

# Calculate shares
compare_sales <- compare_sales %>% 
  mutate(gp_sale=sales_million_m_wh*1000000) %>%
  mutate(share=round(gp_sale/total*100,2)) %>%
  select(year, share)
compare_sales
```

# Capacity additions by state and by year

Next, we can calculate annual renewable energy capacity addition using the dataset for existing capacity. We can use `lag()` function to create lag values to calculate the annual difference. 

```{r}
# Getting the data types right
net_capa_raw$nameplate_capacity_megawatts <- as.numeric(gsub(",","",net_capa_raw$nameplate_capacity_megawatts))

net_capa_tot <- net_capa_raw %>% filter(year>=2009, producer_type=="Total Electric Power Industry",
                                        fuel_source %in% c("Wind", "Solar Thermal and Photovoltaic", "Geothermal")) %>% 
  select(year, state_code, fuel_source, capacity = nameplate_capacity_megawatts) %>%
  group_by(year, state_code) %>%
  summarize(capacity=sum(capacity))
net_capa_tot
```

# Correlation between green power sales and RE capacity by US state

```{r}
net_capa_2019 <- net_capa_tot %>% filter(year==2019) %>% rename(state=state_code)
re_go_2019 <- merge(net_capa_2019, gp_2019_gen_raw, by=c("state"))
cor(re_go_2019$capacity, re_go_2019$total)

ggplot(re_go_2019, aes(x=total, y=capacity)) + geom_point() + geom_smooth() +  
  theme_minimal() +
  labs(title= "Green power sales vs. RE capacity by US state",
       x="Green power sales (MWh)", y="RE Capacity")
```

# Comparison of green power sales and RE additions 2010-2019 (US total)

```{r}
net_capa_us <- net_capa_tot %>% group_by(year) %>%
  summarize(capacity_re = sum(capacity)) %>%
  mutate(cap_lag = lag(capacity_re),
         cap_add = capacity_re-cap_lag) %>%
  filter(year >= 2010)

re_gp_us <- merge(net_capa_us, gp_historical_raw, by="year") 

cor(re_gp_us$capacity_re, re_gp_us$sales_million_m_wh)

ggplot(re_gp_us, aes(x=sales_million_m_wh, y=capacity_re)) +
  geom_point() + geom_smooth() + 
  labs(title= "Relationship between 2019 Green power sales vs. RE capacity by year",
       x="Green power sales (milion MWh)", y="RE Capacity (MW)") 

ggplot(re_gp_us, aes(x=sales_million_m_wh, y=cap_add)) +
  geom_point() + geom_smooth() + 
  labs(title= "Relationship between US Green power sales vs. Annual RE capacity addition by year",
       x="Green power sales (milion MWh)", y="Annual RE Capacity addition (MW)") + theme_void()

```

